Title Text

### 2013-05-03 papers/gpce2013/gpce2013.tex
Slight change in the intro

(TODO more context)

Writing large Web applications is known to be
challenging~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity} mainly because the business logic is
distributed into heterogeneous client-side and server-side
environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}. This makes it difficult to share code
between client-side and server-side (FIXME more on the motivation for code sharing?).

Using the same programming language on both server-side and client-side could reduce this gap, however the JavaScript
language -- which is currently the only action language natively supported by all Web browsers -- has several
drawbacks making it hardly suitable for large code bases (\eg no static typing, no module system, verbose syntax,
\etc). An increasing number of programming languages support generating JavaScript (\eg GWT,
SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}}, Dart,
Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}},
ClojureScript~\cite{McGranaghan11_ClojureScript}, Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}},
Haxe~\cite{Cannasse08_HaXe} or Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}), increasing the panel
of programming languages available to write both the client-side and the server-side code of a Web application.

Having a full featured, cutting edge, programming language that can be run on both client-side and server-side can
help developers to write more maintainable code, however by abstracting over the differences of the client-side and
the server-side environments, the code may suffer from performance issues. Performance is a primary concern in Web
applications, because they are supposed to run on a broad range of devices, from the powerful desktop personal
computer to the less powerful smartphone. “every 100~ms delay costs 1\% of sales”, said Amazon in 2006.

>>>>

Rich Internet Applications have interesting properties: they require no installation or deployment on
clients and enables large scale collaborative experiences. However, writing large Web applications is known to be
difficult~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}. One challenge comes from the fact
that the business logic is scattered into heterogeneous client-side and server-side
environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}. For instance, HTML fragments may be
built from the server-side when a page is requested by a client, but they may also be built from the client-side to
perform an incremental update subsequent to a user action. How could developers write HTML fragment definitions once
and render them on both client-side and server-side? The more interactive the application is, the more logic needs to
be duplicated between the server-side and the client-side.

Using the same programming language on both server-side and client-side could reduce the gap between client-side and
server-side environments, however the JavaScript language -- which is currently the only action language natively
supported by all Web browsers -- has several drawbacks making it hardly suitable for large code bases (\eg no static
typing, no module system, verbose syntax, \etc). An increasing number of programming languages support generating
JavaScript (\eg GWT, SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}}, Dart,
Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}},
ClojureScript~\cite{McGranaghan11_ClojureScript}, Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}},
Haxe~\cite{Cannasse08_HaXe} or Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}), increasing the panel
of programming languages available to write both the client-side and the server-side code of a Web application.

Having a full featured, cutting edge, programming language that runs on both client-side and server-side can help
developers to write more maintainable code, however by abstracting over the differences of the client-side and the
server-side environments, the code may suffer from performance issues. Performance is a primary concern in Web
applications, because they are supposed to run on a broad range of devices, from the powerful desktop personal
computer to the less powerful smartphone. “Every 100~ms delay costs 1\% of sales”, said Amazon in 2006.
